Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-29 paper.tex
Make data tables systematic

\begin{table}
\caption{Summary for VCF Zarr conversion of 
Genomics England WGS data for chromosome 2 (78,195 samples, 59,880,903 variants),
consisting of 44 fields and 2.54 TiB of storage ($\sim$5X compression over
source gzipped VCF).
Each field is stored independently
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes
(omitting values < 0.01\%).
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_AD & int16 & 658.4G & 26 & 25.35\% \\
/call\_GQ & int16 & 654.5G & 13 & 25.20\% \\
/call\_DP & int16 & 570.0G & 15 & 21.95\% \\
/call\_DPF & int16 & 447.1G & 20 & 17.22\% \\
/call\_PL & int16 & 162.6G & 160 & 6.26\% \\
/call\_GQX & int16 & 41.0G & 210 & 1.58\% \\
/call\_FT & string & 25.0G & 1400& 0.96\% \\
/call\_genotype & int8 & 21.5G & 410& 0.83\% \\
/call\_genotype\_mask & bool & 12.8G & 680& 0.49\% \\
/call\_genotype\_phased & bool & 2.4G & 1900& 0.09\% \\
/call\_PS & int8 & 383.4M & 12000& 0.01\% \\
% /call\_ADF & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADR & int8 & 383.4M & 12000.0 & 0.01\% \\
/variant\_position & int32 & 111.6M & 2 & \\
/variant\_quality & float32 & 87.4M & 2.6 & \\
/variant\_allele & string & 69.3M & 13 & \\
% /variant\_OLD\_MULTIALLELIC & object & 55.88M & 8.2 & 0.00\% \\
/variant\_AN & int32 & 47.3M & 4.8 & \\
% /variant\_AC & int32 & 47M & 4.9 & 0.00\% \\
% /variant\_AC\_Het & int32 & 46.62M & 4.9 & 0.00\% \\
% /variant\_ABratio & float32 & 20.23M & 11.0 & 0.00\% \\
% /variant\_MendelSite & float32 & 20.09M & 11.0 & 0.00\% \\
% /variant\_medianGQ & float32 & 19.16M & 12.0 & 0.00\% \\
% /variant\_completeGTRatio & float32 & 14.98M & 15.0 & 0.00\% \\
% /variant\_phwe\_eur & float32 & 14.46M & 16.0 & 0.00\% \\
% /variant\_phwe\_afr & float32 & 13.37M & 17.0 & 0.00\% \\
% /variant\_medianDepthNonMiss & float32 & 12.17M & 19.0 & 0.00\% \\
% /variant\_medianDepthAll & float32 & 11.99M & 19.0 & 0.00\% \\
% /variant\_phwe\_sas & float32 & 11.59M & 20.0 & 0.00\% \\
% /variant\_AC\_Hom & int32 & 11.12M & 21.0 & 0.00\% \\
% /variant\_missingness & float32 & 9.3M & 25.0 & 0.00\% \\
% /variant\_phwe\_eas & float32 & 7.12M & 32.0 & 0.00\% \\
/variant\_filter & bool & 6.4M & 570 & \\
% /variant\_AC\_Hemi & int32 & 849.43 KiB & 280.0 & 0.00\% \\
% /variant\_composite\_filter & bool & 433.38 KiB & 130.0 & 0.00\% \\
% /variant\_OLD\_CLUMPED & object & 384.18 KiB & 1200.0 & 0.00\% \\
% /variant\_id & object & 298.76 KiB & 1600.0 & 0.00\% \\
% /variant\_phwe\_amr & float32 & 269.52 KiB & 870.0 & 0.00\% \\
% /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
/sample\_id & str & 268.1K & 2.3 & \\
% /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
% /contig\_length & int64 & 3.79 KiB & 5.3 & 0.00\% \\
% /contig\_id & object & 2.3 KiB & 8.8 & 0.00\% \\
% /filter\_id & object & 808 bytes & 0.6 & 0.00\% \\
%%%% OLD for chr20
% call\_AD &  int16 & 179.7G & 26 & 24.0\\
% call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
% call\_DP &  int16 & 141.8G & 16 & 18.9 \\
% call\_DPF& int16  & 115.1G & 20 & 15.3\\
% call\_FT &  string & 58.5G & 160 & 7.8 \\
% call\_PL &  int16 & 51.4G & 140 & 6.9 \\
% call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
% call\_genotype & int8 & 6.1G & 380 & 0.8 \\
% call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
% call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
% call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
% variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
% variant\_allele & string & 21.5M & 11 \\
% variant\_position & int32 & 12.7M & 4.7 \\
% variant\_ABratio & float32 & 9.6M & 6.3 \\
% variant\_AN & int32 & 9.6M & 6.3 \\
% variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of 
Our Future Health genotype data for chromosome 22 (651,050 samples, 10,221
variants) consisting of 19 fields and 17 GiB of storage ($\sim2.5$X compression
over source gzipped VCF).
% 42.3  / 17 = 2.4882
Here we use a chunk size of 1,000 variants $\times$ 10,000 samples and 
apply the BitRound(5) filter to the large floating point fields (see text for 
details). Shown are the top 7 fields consuming at 0.001\% or more of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
\label{tab-ofh-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_LRR & float32 & 8.35 GiB & 3.0 & 52.08\% \\
call\_BAF & float32 & 6.02 GiB & 4.1 & 37.55\% \\
call\_GS & float32 & 961.37 MiB & 26.0 & 5.86\% \\
call\_genotype & int8 & 657.21 MiB & 19.0 & 4.00\% \\
call\_genotype\_mask & bool & 79.48 MiB & 160.0 & 0.48\% \\
sample\_id & str & 2.93 MiB & 1.7 & 0.02\% \\
call\_genotype\_phased & bool & 644.72 KiB & 10000.0 & 0.00\% \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of 
All of Us exome-like data genotype data for chromosome 20 (245,394 samples, 
715,256 variants) 
consisting of 25 fields and 8.5 GiB of storage ($\sim1.1$X larger
% 8.5 / 7.44 = 1.142
than source gzipped VCF; see text for discussion).
Here we use the local alleles fields \texttt{call\_LA} and \texttt{call\_LAD}; 
see text for details.
For column details, see Table~\ref{tab-genomics-england-data}.
\label{tab-aou-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
/call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
/call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
/call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
/call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
/call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
/call\_FT & object & 498.61 MiB & 2700.0 & 6.00\% \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of the Norway Spruce
%TODO The variants number here does not match the text "3.8 million"
WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
consisting of 34 fields and 2.3 TiB of storage ($\sim 3.2$X smaller 
than source gzipped VCFs). The chunk size is 10,000 variants $\times$
1,063 samples (i.e., a single sample chunk). This uses local 
alleles to reduce the 
space required by the PL field; see the text for discussion.
For column details see the caption of
Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}

\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_LPL & int16 & 1.26 T & 17.0 & 54.76\% \\
/call\_LA & int8 & 415.39 G & 18.0 & 17.63\% \\
/call\_genotype & int8 & 381.28 G & 19.0 & 16.18\% \\
/call\_genotype\_mask & bool & 103.84 G & 71.0 & 4.41\% \\
/variant\_DP4 & int32 & 26.24 G & 2.1 & 1.11\% \\
/variant\_MQSBZ & float32 & 13.08 G & 1.1 & 0.56\% \\
/variant\_RPBZ & float32 & 13.04 G & 1.1 & 0.55\% \\
/variant\_MQBZ & float32 & 12.97 G & 1.1 & 0.55\% \\
/variant\_SCBZ & float32 & 12.94 G & 1.1 & 0.55\% \\
/variant\_VDB & float32 & 12.84 G & 1.1 & 0.55\% \\
/variant\_quality & float32 & 12.84 G & 1.1 & 0.55\% \\
/variant\_BQBZ & float32 & 12.64 G & 1.1 & 0.54\% \\
/variant\_SGB & float32 & 12.12 G & 1.2 & 0.51\% \\
/variant\_position & int32 & 9.65 G & 1.4 & 0.41\% \\
/variant\_AC & int16 & 7.42 G & 2.8 & 0.31\% \\
/variant\_DP & int32 & 6.71 G & 2.1 & 0.28\% \\
/variant\_allele & object & 5.37 G & 21.0 & 0.23\% \\
/variant\_AN & int16 & 3.05 G & 2.3 & 0.13\% \\
/call\_genotype\_phased & bool & 1.75 G & 2100.0 & 0.07\% \\
/variant\_filter & bool & 1.46 G & 2.4 & 0.06\% \\
/variant\_MQ & int8 & 489.41 M & 7.3 & 0.02\% \\
/variant\_IMF & float32 & 160.95 M & 89.0 & 0.01\% \\
/variant\_IDV & int16 & 120.72 M & 59.0 & 0.01\% \\
% /variant\_length & int16 & 105.19 M & 68.0 & 0.00\% \\
% /variant\_INDEL & bool & 72.18 M & 49.0 & 0.00\% \\
% /variant\_id & object & 34.53 M & 830.0 & 0.00\% \\
% /variant\_id_mask & bool & 32.74 M & 110.0 & 0.00\% \\
% /variant\_contig & int16 & 32.02 M & 220.0 & 0.00\% \\
% /variant\_FS & float32 & 31.67 M & 450.0 & 0.00\% \\
% /variant\_MQ0F & float32 & 31.67 M & 450.0 & 0.00\% \\
% /region\_index & int32 & 3.14 M & 2.7 & 0.00\% \\
% /contig\_length & int64 & 6.77 K & 1.2 & 0.00\% \\
% /contig\_id & object & 6.2 K & 1.3 & 0.00\% \\
% /sample\_id & object & 5.35 K & 1.6 & 0.00\% \\
% /filter\_id & object & 4.43 K & 0.0 & 0.00\% \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\begin{table}
\caption{Summary for VCF Zarr conversion of 
Genomics England WGS data for chromosome 2 (78,195 samples, 59,880,903 variants),
consisting of 44 fields and 2.54 TiB of storage ($\sim$5X compression over
source gzipped VCF).
Each field is stored independently
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes.
Shown are the top 11 fields consuming at least 0.01\% of the overall storage.
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=5.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\%Total} \\
\midrule
call\_AD & int16 & 658.44 GiB & 26.0 & 25.35\% \\
call\_GQ & int16 & 654.45 GiB & 13.0 & 25.20\% \\
call\_DP & int16 & 570.03 GiB & 15.0 & 21.95\% \\
call\_DPF & int16 & 447.09 GiB & 20.0 & 17.22\% \\
call\_PL & int16 & 162.56 GiB & 160.0 & 6.26\% \\
call\_GQX & int16 & 40.99 GiB & 210.0 & 1.58\% \\
call\_FT & str & 25.04 GiB & 1400.0 & 0.96\% \\
call\_genotype & int8 & 21.46 GiB & 410.0 & 0.83\% \\
call\_genotype\_mask & bool & 12.78 GiB & 680.0 & 0.49\% \\
call\_genotype\_phased & bool & 2.35 GiB & 1900.0 & 0.09\% \\
call\_PS & int8 & 383.38 MiB & 12000.0 & 0.01\% \\
call\_ADF & int8 & 383.38 MiB & 12000.0 & 0.01\% \\
call\_ADR & int8 & 383.38 MiB & 12000.0 & 0.01\% \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of 
Our Future Health genotype data for chromosome 22 (651,050 samples, 10,221
variants) consisting of 19 fields and 17 GiB of storage ($\sim2.5$X compression
over source gzipped VCF).
% 42.3  / 17 = 2.4882
Here we use a chunk size of 1,000 variants $\times$ 10,000 samples and 
apply the BitRound(5) filter to the large floating point fields (see text for 
details). 
Shown are the top 6 fields consuming at least 0.01\% of the overall storage 
(see Table~\ref{tab-genomics-england-data} for column details).
\label{tab-ofh-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_LRR & float32 & 8.35 GiB & 3.0 & 52.08\% \\
call\_BAF & float32 & 6.02 GiB & 4.1 & 37.55\% \\
call\_GS & float32 & 961.37 MiB & 26.0 & 5.86\% \\
call\_genotype & int8 & 657.21 MiB & 19.0 & 4.00\% \\
call\_genotype\_mask & bool & 79.48 MiB & 160.0 & 0.48\% \\
sample\_id & str & 2.93 MiB & 1.7 & 0.02\% \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of 
All of Us exome-like data genotype data for chromosome 20 (245,394 samples, 
715,256 variants) 
consisting of 25 fields and 8.5 GiB of storage ($\sim1.1$X larger
% 8.5 / 7.44 = 1.142
than source gzipped VCF; see text for discussion).
Here we use the local alleles fields \texttt{call\_LA} and \texttt{call\_LAD}; 
see text for details.
Shown are the top 14 fields consuming at least 0.01\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
The \texttt{variant\_homozygote\_count} field has been renamed 
\texttt{variant\_hom\_count} for display purposes.
\label{tab-aou-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
call\_FT & str & 498.61 MiB & 2700.0 & 6.00\% \\
call\_genotype\_phased & bool & 17.17 MiB & 9700.0 & 0.21\% \\
variant\_AF & float32 & 5.95 MiB & 43.0 & 0.07\% \\
% /variant\_homozygote\_count & int32 & 4.78 MiB & 54.0 & 0.06\% \\
variant\_hom\_count & int32 & 4.78 MiB & 54.0 & 0.06\% \\
variant\_allele & str & 4.74 MiB & 110.0 & 0.06\% \\
variant\_AC & int32 & 4.24 MiB & 61.0 & 0.05\% \\
variant\_filter & bool & 2.87 MiB & 0.9 & 0.03\% \\
variant\_AN & int32 & 904.75 KiB & 3.1 & 0.01\% \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of the Norway Spruce
%TODO The variants number here does not match the text "3.8 million"
WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
consisting of 34 fields and 2.3 TiB of storage ($\sim 3.2$X smaller 
than source gzipped VCFs). The chunk size is 10,000 variants $\times$
1,063 samples (i.e., a single sample chunk). This uses local 
alleles to reduce the 
space required by the PL field; see the text for discussion.
Shown are the top 21 fields consuming at least 0.01\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
\label{tab-spruce-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_LPL & int16 & 1.26 TiB & 17.0 & 54.76\% \\
call\_LA & int8 & 415.39 GiB & 18.0 & 17.63\% \\
call\_genotype & int8 & 381.28 GiB & 19.0 & 16.18\% \\
call\_genotype\_mask & bool & 103.84 GiB & 71.0 & 4.41\% \\
variant\_DP4 & int32 & 26.24 GiB & 2.1 & 1.11\% \\
variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.56\% \\
variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.55\% \\
variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.55\% \\
variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.55\% \\
variant\_quality & float32 & 12.84 GiB & 1.1 & 0.55\% \\
variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.55\% \\
variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.54\% \\
variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.51\% \\
variant\_position & int32 & 9.65 GiB & 1.4 & 0.41\% \\
variant\_AC & int16 & 7.42 GiB & 2.8 & 0.31\% \\
variant\_DP & int32 & 6.71 GiB & 2.1 & 0.28\% \\
variant\_allele & str & 5.37 GiB & 21.0 & 0.23\% \\
variant\_AN & int16 & 3.05 GiB & 2.3 & 0.13\% \\
call\_genotype\_phased & bool & 1.75 GiB & 2100.0 & 0.07\% \\
variant\_filter & bool & 1.46 GiB & 2.4 & 0.06\% \\
variant\_MQ & int8 & 489.41 MiB & 7.3 & 0.02\% \\
\bottomrule
\end{tabular}
\end{table}
